The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2018-07-12 Paper/jaccardbaetica.tex
add new pic

The aim of this study is to detect the patterns of olive oil production that link amphora workshops and amphoric stamps. Roman provinces such as Baetica became important production and distribution centres during the Roman Empire. However, it remains under debate how this province was organized and whether it is possible to identify patterns in the olive oil market.

Our case of study has been focused on the production processes located in Baetica province (currently Andalusia) from 1st to 3rd AD. In particular, we want to explore economic dynamics that include the production and distribution of olive oil trade. Amphoric stamps are used to identify the presence of different producer groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from different workshops in Baetica province in order to detect a relation between the distribution of amphoric stamps and the economic structure in this province. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

@@

\section{The amphora production in the Roman Empire}

@@

Roman provinces such as Baetica (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, the high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. This product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, thousands of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

\subsection{Quantifying the diversity}

@@

%quizás habría que hablar también del filtrado de códigos que se hizo en python porque antes había 3783 stamps pero se filtró por el número de letras...
\xavi{lo del filtrado lo puedes poner de supplementary information con referencia al antiquity}

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, groupings of similar stamps were not found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

>>>>

The aim of this study is to detect the patterns of olive oil production that link amphora workshops and amphoric stamps. Our case of study has been focused on the production processes located in Baetica province (currently Andalusia) from 1st to 3rd AD. In particular, we want to explore economic dynamics that include the production and distribution of olive oil trade. 
Baetica became important production and distribution centres during the Roman Empire. However, it remains under debate how this province was organized and whether it is possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different producer groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from different workshops in Baetica province in order to detect a relation between the distribution of amphoric stamps and the economic structure in this province. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. Given these factors, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places. Roman provinces such as Baetica (currently Andalusia, south Spain) became important production centres of olive oil during the Roman Empire. Olive oil was considered as the liquid gold since it was used in different aspect of the daily life as cooking, hygiene or lighting  \citep{mattingly_d.j._oil_1988}. As a consequence, this product was distributed in large amount of amphorae along the province, mostly to supply the Roman Army and Italy \citep{blazquez_exportacion_1980}.

@@

\section{The Amphoric production in Baetica province}

@@

The high demand of Roman provinces stimulated by the good condition of the Baetica lands allowed to develop a massive infrastructure of olive oil production. The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, thousands of amphora workshops were made to support this high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

@@

\subsection{Quantifying the cultural diversity: Morisita-Horn index}

@@

%quizás habría que hablar también del filtrado de códigos que se hizo en python porque antes había 3783 stamps pero se filtró por el número de letras...
%\xavi{lo del filtrado lo puedes poner de supplementary information con referencia al antiquity}

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, groups of workshops sharing similar stamps were not found in the cluster: the majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.
